---
title: "Megans Deeda RNASeq DESeq Analysis"
author: "Tanner Porter"
date: "2023-12-07"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---



```{r setup, include=FALSE}
#loadpackages
library(tidyverse)
library(DESeq2)
```

# Basic Manipulation of the Data
Import the count table (sequencing data), metadata, pair them and create a DESeq object.

```{r Data Import}

#Read in the sequencing data...
TB_RNA <- as.matrix(read.csv("/scratch/mfolkerts/Deeda_RNA_Seq/Latest_data/countTable_dedup_nocontrols.txt", sep="\t", row.names = "Geneid"))

#Read in the metadata file.
Metadata <- readxl::read_xlsx("/scratch/tporter/TB_20231117_Megans_DESeq/TGEN CASS Shipment 1 Metadata_07-02-2024.xlsx")

Metadata_2 <- read_csv("/scratch/mfolkerts/Deeda_RNA_Seq/Latest_data/metadata_RNASeq_Deeda.csv")

Metadata <- left_join(Metadata, Metadata_2)

rm(Metadata_2)

# row.names(Metadata) <- Metadata$`Participant ID` #Depreciated

Metadata <- Metadata %>%
  mutate(Participant_ID2 = `Participant ID`) %>%
  column_to_rownames(var = "Participant_ID2")

#Convert metadata to factors
Metadata$Group <- factor(str_replace(Metadata$Grouping, " ", "_"), levels = c("CONTROL_2", "CONTROL_1", "CASES")) #Remove was from names

Metadata$Run <- factor(Metadata$Run)

Metadata$Gender <- factor(Metadata$Gender)

Metadata$HIV_Status <- factor(Metadata$`HIV Status`)

Metadata$CASS_Smear <- factor(Metadata$`CASS Smear`)

Metadata$CASS_Culture <- factor(Metadata$`CASS Culture`)

Metadata$Simple_Group <- ifelse(grepl("CASES",Metadata$Group), "CASES", "CONTROL")

#Convert metadata to factors
Metadata$Simple_Group <- factor(Metadata$Simple_Group, levels = c("CONTROL", "CASES")) #Remove was from names

Metadata$Simple_Group


########
########
# Removes 1 sample that was not sequenced.
########
########
Metadata <- Metadata %>%
  filter(row.names(Metadata) %in% colnames(TB_RNA))
########
########


#See if there are issues between the dataset.
rownames(Metadata) %in% colnames(TB_RNA)
#reorder the columns in metatdata and sampledata so that they are in the same order for DeSeq2
TB_RNA <- TB_RNA[, rownames(Metadata)]
#check if it worked
all(rownames(Metadata) == colnames(TB_RNA))


##### Since there is a bunch of metadata missing...
Metadata_CASS_SMEAR <- Metadata %>%
  filter(!is.na(`CASS Smear`))



TB_RNA_CASS_SMEAR <- TB_RNA[,(colnames(TB_RNA) %in% rownames(Metadata_CASS_SMEAR)), drop = F]
#check if it worked
all(rownames(Metadata_CASS_SMEAR) == colnames(TB_RNA_CASS_SMEAR)) 




```
# Join with annotations file
```{r}
genes.anno <- fst::read_fst("/scratch/tporter/TB_20231117_Megans_DESeq/GRCh38_GENCODE40LNCipedia_geneAnnotations.fst") %>%
  dplyr::select(gene_id, gene_type, gene_name, biotype_class) %>%
  unique()

```

# General Data Exploration and Visualization
```{r}
Metadata <- Metadata %>%
  mutate(Name = row.names(Metadata))

Metadata %>%  
  ggplot(aes(y = Read_count, x = Name, fill = Group)) +
    geom_bar(stat = "identity")+
    theme_bw()+
    scale_y_log10()+
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+
    ggsci::scale_fill_npg() +
    ggtitle("Reads across samples and groups")

Metadata %>%  
  ggplot(aes(y = Read_count, x = Group, fill = Group)) +
    geom_boxplot()+
    theme_bw()+
    scale_y_log10()+
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+
    ggsci::scale_fill_npg()+
    ggtitle("Reads across samples and groups")


Metadata <- full_join(Metadata, data_frame(Name = colnames(TB_RNA),
                   N_genes = TB_RNA %>%colSums()))

TB_RNA_DF <- as_data_frame(TB_RNA)

TB_RNA_DF$Gene <- row.names(TB_RNA)

TB_RNA_DF <- TB_RNA_DF %>%
  pivot_longer(!Gene, values_to = "Reads", names_to = "Name")

TB_RNA_DF <- left_join(TB_RNA_DF, Metadata)

TB_RNA_DF %>%
  filter(Reads > 0) %>%
  group_by(Name, Group) %>%
  tally(name = "N_Genes_0") %>%
  ggplot(aes(x = Name, y = N_Genes_0, fill = Group))+
    geom_bar(stat = "identity")+
    theme_bw()+
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+
    ggsci::scale_fill_npg()+
    ggtitle("n genes with reads across groups")

TB_RNA_DF %>%
  filter(Reads > 0) %>%
  group_by(Name, Group) %>%
  tally(name = "N_Genes_0") %>%
  ggplot(aes(x = Group, y = N_Genes_0, fill = Group))+
    geom_boxplot()+
    theme_bw()+
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+
    ggsci::scale_fill_npg()+
    ylab("n Genes Identified (reads > 0)")

TB_RNA_DF %>%
  filter(Reads > 0) %>%
  group_by(Name, Group, Read_count) %>%
  tally(name = "N_Genes_0") %>%
  ggplot(aes(x = Read_count, y = N_Genes_0, col = Group))+
    geom_point()+
    theme_bw()+
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+
    ggsci::scale_color_npg()+
    scale_x_log10() +
    ggtitle("n genes with reads vs read count across groups")

TB_RNA_DF %>%
  filter(Reads > 0) %>%
  group_by(Name) %>%
  tally(name = "N_Genes_0") %>%
  summarise(median = median(N_Genes_0, col = Group))

Metadata <- left_join(Metadata,
          TB_RNA_DF %>%
            filter(Reads > 0) %>%
            group_by(Name) %>%
            tally(name = "N_Genes_0")) %>%
            mutate(Gene_Count_Group = ifelse(N_Genes_0 > 25730, "High", "Low"))

Metadata$CASS_outcome <- factor(Metadata$`CASS outcome`, levels = c("Negative", "Positive")) #Remove was from names

Metadata$Smear_Group <- factor(Metadata$Smear, levels = c("Negative", "Positive")) #Remove was from names
```



# Conduct DESeq Analysis for Group
```{r}
#This creates a DESeq object
DeedaRNASeq_Groups <- DESeqDataSetFromMatrix(countData = TB_RNA,
                              colData = Metadata,
                              design = ~ Group + Run + Gender + HIV_Status)

DeedaRNASeq_Groups

#Conduct RNASeq Analysis for full control
DeedaRNASeq_Analysis_Groups <- DESeq(DeedaRNASeq_Groups, test="Wald")

plotDispEsts(DeedaRNASeq_Analysis_Groups)

resultsNames(DeedaRNASeq_Analysis_Groups)

DeedaRNASeq_Analysis_Groups@design
```

# Conduct DESeq Analysis for CASS_outcome
```{r}
#This creates a DESeq object
DeedaRNASeq_CASS_outcome <- DESeqDataSetFromMatrix(countData = TB_RNA,
                              colData = Metadata,
                              design = ~ CASS_outcome + Run + Gender + HIV_Status)

DeedaRNASeq_CASS_outcome

#Conduct RNASeq Analysis for full control
DeedaRNASeq_Analysis_CASS_outcome <- DESeq(DeedaRNASeq_CASS_outcome, test="Wald")

plotDispEsts(DeedaRNASeq_Analysis_CASS_outcome)
  
resultsNames(DeedaRNASeq_Analysis_CASS_outcome)

DeedaRNASeq_Analysis_CASS_outcome@design
```

# Conduct DESeq Analysis for CAAS Smear
```{r}
#This creates a DESeq object
DeedaRNASeq_Smear <- DESeqDataSetFromMatrix(countData = TB_RNA_CASS_SMEAR,
                              colData = Metadata_CASS_SMEAR,
                              design = ~ CASS_Smear + Run + Gender + HIV_Status)

DeedaRNASeq_Smear

#Conduct RNASeq Analysis for full control
DeedaRNASeq_Analysis_Smear <- DESeq(DeedaRNASeq_Smear, test="Wald")

plotDispEsts(DeedaRNASeq_Analysis_Smear)
  
resultsNames(DeedaRNASeq_Analysis_Smear)

DeedaRNASeq_Analysis_Smear@design
```

# Conduct DESeq Analysis for CASS Smear
```{r}
#This creates a DESeq object
DeedaRNASeq_Culture <- DESeqDataSetFromMatrix(countData = TB_RNA_CASS_SMEAR,
                              colData = Metadata_CASS_SMEAR,
                              design = ~ CASS_Culture + Run + Gender + HIV_Status)

DeedaRNASeq_Culture

#Conduct RNASeq Analysis for full control
DeedaRNASeq_Analysis_Culture <- DESeq(DeedaRNASeq_Culture, test="Wald")

plotDispEsts(DeedaRNASeq_Analysis_Culture)
  
resultsNames(DeedaRNASeq_Analysis_Culture)

DeedaRNASeq_Analysis_Culture@design
```

# Contrast Analysis
```{r}

### Group Contrast Analysis
Contrast_Groups_CONTROL1_CASES <- data.frame(results(DeedaRNASeq_Analysis_Groups, contrast=c("Group","CONTROL_1","CASES"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "Group- Control_1 vs. Cases")%>%
  mutate(gene_id = row.names(.))

Contrast_Groups_CONTROL2_CASES <- data.frame(results(DeedaRNASeq_Analysis_Groups, contrast=c("Group","CONTROL_2","CASES"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "Group- Control_2 vs. Cases")%>%
  mutate(gene_id = row.names(.))

Contrast_Groups_CONTROL2_CONTROL1 <- data.frame(results(DeedaRNASeq_Analysis_Groups, contrast=c("Group","CONTROL_2","CONTROL_1"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "Group- Control_2 vs. Control_1")%>%
  mutate(gene_id = row.names(.))

Contrast_Groups_Run<- data.frame(results(DeedaRNASeq_Analysis_Groups, contrast=c("Run","1", "2"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "Group- Run 1 vs. 2") %>%
  mutate(gene_id = row.names(.))

Contrast_Groups_Gender<- data.frame(results(DeedaRNASeq_Analysis_Groups, contrast=c("Gender","Male", "Female"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "Group- Gender Male vs. Female") %>%
  mutate(gene_id = row.names(.))

Contrast_Groups_HIV<- data.frame(results(DeedaRNASeq_Analysis_Groups, contrast=c("HIV_Status","Positive", "Negative"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "Group- HIV Positive vs. Negative") %>%
  mutate(gene_id = row.names(.))

### CASS_outcome Contrast Analysis

DeedaRNASeq_Analysis_CASS_outcome@design

Contrast_CASS_outcome <- data.frame(results(DeedaRNASeq_Analysis_CASS_outcome, contrast=c("CASS_outcome","Negative", "Positive"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Outcome- Negative vs. Positive") %>%
  mutate(gene_id = row.names(.))

Contrast_CASS_Run<- data.frame(results(DeedaRNASeq_Analysis_CASS_outcome, contrast=c("Run","1", "2"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Outcome- Run 1 vs. 2") %>%
  mutate(gene_id = row.names(.))

Contrast_CASS_Gender<- data.frame(results(DeedaRNASeq_Analysis_CASS_outcome, contrast=c("Gender","Male", "Female"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Outcome- Gender Male vs. Female") %>%
  mutate(gene_id = row.names(.))

Contrast_CASS_HIV<- data.frame(results(DeedaRNASeq_Analysis_CASS_outcome, contrast=c("HIV_Status","Positive", "Negative"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Outcome- HIV Positive vs. Negative") %>%
  mutate(gene_id = row.names(.))

DeedaRNASeq_Analysis_CASS_outcome@design

### Smear_Group Contrast Analysis

DeedaRNASeq_Analysis_Smear@design

Contrast_Smear_outcome <- data.frame(results(DeedaRNASeq_Analysis_Smear, contrast=c("CASS_Smear","Negative", "Positive"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Smear- Negative vs. Positive") %>%
  mutate(gene_id = row.names(.))

Contrast_Smear_Run<- data.frame(results(DeedaRNASeq_Analysis_Smear, contrast=c("Run","1", "2"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Smear- Run 1 vs. 2") %>%
  mutate(gene_id = row.names(.))

Contrast_Smear_Gender<- data.frame(results(DeedaRNASeq_Analysis_Smear, contrast=c("Gender","Male", "Female"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Smear- Gender Male vs. Female") %>%
  mutate(gene_id = row.names(.))

Contrast_Smear_HIV<- data.frame(results(DeedaRNASeq_Analysis_Smear, contrast=c("HIV_Status","Positive", "Negative"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Smear- HIV Positive vs. Negative") %>%
  mutate(gene_id = row.names(.))

### CASS CultureGroup Contrast Analysis

DeedaRNASeq_Analysis_Culture@design

Contrast_Culture_outcome <- data.frame(results(DeedaRNASeq_Analysis_Culture, contrast=c("CASS_Culture","Negative", "Positive"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Culture- Negative vs. Positive") %>%
  mutate(gene_id = row.names(.))

Contrast_Culture_Run<- data.frame(results(DeedaRNASeq_Analysis_Culture, contrast=c("Run","1", "2"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Culture- Run 1 vs. 2") %>%
  mutate(gene_id = row.names(.))

Contrast_Culture_Gender<- data.frame(results(DeedaRNASeq_Analysis_Culture, contrast=c("Gender","Male", "Female"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Culture- Gender Male vs. Female") %>%
  mutate(gene_id = row.names(.))

Contrast_Culture_HIV<- data.frame(results(DeedaRNASeq_Analysis_Culture, contrast=c("HIV_Status","Positive", "Negative"), alpha=0.05)) %>%
  filter(padj < 0.05) %>%
  mutate(Contrast = "CASS Culture- HIV Positive vs. Negative") %>%
  mutate(gene_id = row.names(.))
```

# Results Combining

```{r}
# List all objects in the workspace
all_objects <- ls()

# Filter objects that begin with "Contrast_"
contrast_objects <- grep("^Contrast_", all_objects, value = TRUE)

contrast_objects

# Check if there are any matching objects
if (length(contrast_objects) > 0) {
  # Combine (bind) the objects using do.call and rbind
  Out <- do.call(rbind, lapply(contrast_objects, get))
  
  # Print or use the combined_contrasts object as needed
  head(Out)
} else {
  print("No objects with the prefix 'Contrast_' found.")
}

head(Out)

Out <- right_join(genes.anno, Out)

```

# BioMart
```{r}
require('biomaRt')

mart <- useMart('ENSEMBL_MART_ENSEMBL')
mart <- useDataset('hsapiens_gene_ensembl', mart)

# annotLookup <- getBM(
#   mart = mart,
#   attributes = c(
#     'ensembl_gene_id',
#     'gene_biotype',
#     'external_gene_name',
#     'uniprot_gn_symbol',
#     'uniprot_gn_id'),
#   uniqueRows=TRUE)

annotLookup <- getBM(
  mart = mart,
  attributes = c(
    'ensembl_gene_id',
    #'gene_biotype',
    #'external_gene_name',
    'uniprotswissprot'),
  uniqueRows=TRUE)

annotLookup <- annotLookup %>%
  filter(!uniprotswissprot == "")

Out$ensembl_gene_id_2 <- Out$gene_id

Out <- Out %>%
  separate(ensembl_gene_id_2, into = c("ensembl_gene_id", "dot"), sep = "\\.")

Out_New <- left_join(Out, annotLookup, by = c("ensembl_gene_id"))

#Out_New <- Out_New[,c(17, 1:16)]

write.csv(Out_New, "/scratch/tporter/TB_20231117_Megans_DESeq/Deeda_RNA_Seq_Compiled_Contrast_Results_with_SWISSUNIPROT_Final.csv")

# 1 duplicated uniprotswiss id which has 2 proteins
Out_New %>%
  group_by(Contrast, gene_type, uniprotswissprot) %>%
  tally() %>%
  filter() %>%
  arrange(desc(n)) %>%
  filter(!is.na(uniprotswissprot))


Dups <- Out_New %>%
  group_by(Contrast, gene_type, ensembl_gene_id) %>%
  tally() %>%
  filter() %>%
  filter(n > 1) %>%
  ungroup() %>%
  dplyr::select(ensembl_gene_id) %>%
  distinct()

annotLookup %>%
  filter(ensembl_gene_id %in% Dups$ensembl_gene_id)


#write.csv(Check, "/scratch/tporter/TB_20231117_Megans_DESeq/Deeda_RNA_Seq_Compiled_Contrast_Results_with_SWISSUNIPROT_RedundancyCheck.csv")
```

# Read Count Visualizations

```{r}
library("pheatmap")

select <- order(rowMeans(counts(DeedaRNASeq_Analysis_CASS_outcome, normalized=TRUE)),
                decreasing=TRUE)[1:25]

df <- as.data.frame(colData(DeedaRNASeq_Analysis_Groups)[,c("Group", "CASS_outcome", "Smear_Group","Gender","Run","HIV_Status")])

df <- as.data.frame(colData(DeedaRNASeq_Analysis_Groups)[,c("Group", "Gender", "CASS_outcome")])

ntd <- normTransform(DeedaRNASeq_Analysis_CASS_outcome)

pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=FALSE, annotation_col=df)

vsd <- vst(DeedaRNASeq_Analysis_Groups, blind=FALSE)

pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df)

plotPCA(vsd, intgroup=c("Gene_Count_Group"))
plotPCA(vsd, intgroup=c("Group", "Gender"))
plotPCA(vsd, intgroup=c("Run"))
plotPCA(vsd, intgroup=c("Group"))
plotPCA(vsd, intgroup=c("Gender"))

plotPCA(ntd, intgroup=c("Gene_Count_Group"))
plotPCA(ntd, intgroup=c("Group", "Run"))
plotPCA(ntd, intgroup=c("Run"))
plotPCA(ntd, intgroup=c("Group"))

```

```{r}
Plotly <- Out %>%
  filter(gene_type == "protein_coding") %>%
  mutate(Contrast_Group = ifelse(grepl("HIV", Contrast), "HIV",
                                 ifelse(grepl("Gender", Contrast), "Gender",
                                        ifelse(grepl("Run", Contrast), "Run", "Experimental")))) %>%
  ggplot(aes(x = gene_name, y = Contrast, fill = log2FoldChange))+
    geom_tile()+
    theme_bw()+
    facet_wrap(~Contrast_Group, scales = "free", ncol = 1)+
    theme(axis.text.x = element_text(angle = 90, hjust = 1))+
    scale_fill_gradientn(colours = c("blue4", "green4", "orange", "red", "red", "red"))

Plotly

plotly::ggplotly(Plotly)

htmlwidgets::saveWidget(plotly::as_widget(plotly::ggplotly(Plotly, width = 10000, height = 2500)),
                        file="/scratch/tporter/TB_20231117_Megans_DESeq/Contrast_Figure_Across_Groups.html", selfcontained = TRUE)
```

# General PCA on all genes

```{r}
#Create New DeSeq Obhect Using this for reference file:///Users/tporter/Downloads/RNASEQ20_Day3_HandsOn%20(1).pdf
PCA_Data <- DESeqDataSetFromMatrix(countData = TB_RNA,
                              colData = Metadata,
                              design = ~ Group + Run + Gender + HIV_Status)

PCA_Data <- estimateSizeFactors(PCA_Data)

sizeFactors(PCA_Data)

PCA_normalized <- rlog(PCA_Data, blind = T)

plotPCA(PCA_normalized, intgroup=c("Gender")) +
    theme_bw() +
    stat_ellipse(level = 0.95) +
    ggtitle("Gender")+
    ggsci::scale_color_npg()

plotPCA(PCA_normalized, intgroup=c("Group")) +
    theme_bw() +
    stat_ellipse(level = 0.95) +
    ggtitle("Group")+
    ggsci::scale_color_npg()

plotPCA(PCA_normalized, intgroup=c("Run")) +
    theme_bw() +
    stat_ellipse(level = 0.95) +
    ggtitle("Run")+
    ggsci::scale_color_npg()

plotPCA(PCA_normalized, intgroup=c("HIV_Status")) +
    theme_bw() +
    stat_ellipse(level = 0.95) +
    ggtitle("HIV Status")+
    ggsci::scale_color_npg()
```

# Group Variable PCA
```{r}
Group <- Out_New %>%
  filter(grepl("Group- Control", Contrast))

Filter <- Group$gene_id

Filter

#Create New DeSeq Obhect Using this for reference file:///Users/tporter/Downloads/RNASEQ20_Day3_HandsOn%20(1).pdf
PCA_Data <- DESeqDataSetFromMatrix(countData = TB_RNA,
                              colData = Metadata,
                              design = ~ Group + Run + Gender + HIV_Status)

PCA_Data <- estimateSizeFactors(PCA_Data)

sizeFactors(PCA_Data)

PCA_normalized <- vst(PCA_Data, blind = T)

# Convert tibble to data frame
PCA_normalized <- as.data.frame(head(assay(PCA_normalized), nrow(PCA_normalized)))

PCA_normalized$gene <- rownames(PCA_normalized)

PCA_normalized_Filtered <- PCA_normalized %>%
  filter(gene %in% Filter)

PCA_normalized_Filtered <- PCA_normalized_Filtered %>%
  pivot_longer(-gene, values_to = "Norm_Count", names_to = "Name") %>%
  pivot_wider(values_from = Norm_Count, names_from = gene)

PCA <- prcomp(PCA_normalized_Filtered[,-c(1)], center = T, scale = T)

PCA_Table <- data_frame("Name" = PCA_normalized_Filtered$Name, "PC1" = PCA$x[,1], "PC2" = PCA$x[,2])

PCA_Table

names(PCA_Table)

Temp <- right_join(PCA_Table,
           dplyr::select(Metadata, `Participant ID`, Group, Run, Gender, HIV_Status) %>%
             pivot_longer(Group:HIV_Status, names_to = "Variable", values_to = "Group"),
           by = c("Name" = "Participant ID"))

Temp %>%
  ggplot(aes(x = PC1, y = PC2, col = Group))+
    geom_point()+
    stat_ellipse(level = 0.95)+
    ggsci::scale_color_npg()+
    theme_bw()+
    facet_wrap(~Variable)+
    ggtitle("Principle Component Analysis Across Normalized Read Counts of Significant `Group` Genes PC1: 31.4%, PC2: 9.5%")

Temp %>%
  filter(Variable == "Group") %>%
  ggplot(aes(x = PC1, y = PC2, col = Group))+
    geom_point()+
    stat_ellipse(level = 0.95)+
    ggsci::scale_color_npg()+
    theme_bw()+
    facet_wrap(~Variable)+
    ggtitle("Principle Component Analysis Across Normalized Read Counts of Significant `Group` Genes PC1: 31.4%, PC2: 9.5%")

summary(PCA)
```

# HIV Variable PCA
```{r}
Group <- Out_New %>%
  filter(grepl("Group- HIV", Contrast))

Filter <- Group$gene_id

Filter

#Create New DeSeq Obhect Using this for reference file:///Users/tporter/Downloads/RNASEQ20_Day3_HandsOn%20(1).pdf
PCA_Data <- DESeqDataSetFromMatrix(countData = TB_RNA,
                              colData = Metadata,
                              design = ~ Group + Run + Gender + HIV_Status)

PCA_Data <- estimateSizeFactors(PCA_Data)

sizeFactors(PCA_Data)

PCA_normalized <- vst(PCA_Data, blind = T)

# Convert tibble to data frame
PCA_normalized <- as.data.frame(head(assay(PCA_normalized), nrow(PCA_normalized)))

PCA_normalized$gene <- rownames(PCA_normalized)

PCA_normalized_Filtered <- PCA_normalized %>%
  filter(gene %in% Filter)

PCA_normalized_Filtered <- PCA_normalized_Filtered %>%
  pivot_longer(-gene, values_to = "Norm_Count", names_to = "Name") %>%
  pivot_wider(values_from = Norm_Count, names_from = gene)

PCA <- prcomp(PCA_normalized_Filtered[,-c(1)], center = T, scale = T)

PCA_Table <- data_frame("Name" = PCA_normalized_Filtered$Name, "PC1" = PCA$x[,1], "PC2" = PCA$x[,2])

PCA_Table

names(PCA_Table)

Temp <- right_join(PCA_Table,
           dplyr::select(Metadata, `Participant ID`, Group, Run, Gender, HIV_Status) %>%
             pivot_longer(Group:HIV_Status, names_to = "Variable", values_to = "Group"),
           by = c("Name" = "Participant ID"))

Temp %>%
  ggplot(aes(x = PC1, y = PC2, col = Group))+
    geom_point()+
    stat_ellipse(level = 0.95)+
    ggsci::scale_color_npg()+
    theme_bw()+
    facet_wrap(~Variable)+
    ggtitle("Principle Component Analysis Across Normalized Read Counts of Significant HIV Genes PC1: 27%, PC2: 12%")

summary(PCA)
```

# Gender Variable PCA
```{r}
Group <- Out_New %>%
  filter(grepl("Group- Gender", Contrast))

Filter <- Group$gene_id

Filter

#Create New DeSeq Obhect Using this for reference file:///Users/tporter/Downloads/RNASEQ20_Day3_HandsOn%20(1).pdf
PCA_Data <- DESeqDataSetFromMatrix(countData = TB_RNA,
                              colData = Metadata,
                              design = ~ Group + Run + Gender + HIV_Status)

PCA_Data <- estimateSizeFactors(PCA_Data)

sizeFactors(PCA_Data)

PCA_normalized <- vst(PCA_Data, blind = T)

# Convert tibble to data frame
PCA_normalized <- as.data.frame(head(assay(PCA_normalized), nrow(PCA_normalized)))

PCA_normalized$gene <- rownames(PCA_normalized)

PCA_normalized_Filtered <- PCA_normalized %>%
  filter(gene %in% Filter)

PCA_normalized_Filtered <- PCA_normalized_Filtered %>%
  pivot_longer(-gene, values_to = "Norm_Count", names_to = "Name") %>%
  pivot_wider(values_from = Norm_Count, names_from = gene)

PCA <- prcomp(PCA_normalized_Filtered[,-c(1)], center = T, scale = T)

PCA_Table <- data_frame("Name" = PCA_normalized_Filtered$Name, "PC1" = PCA$x[,1], "PC2" = PCA$x[,2])

PCA_Table

names(PCA_Table)

Temp <- right_join(PCA_Table,
           dplyr::select(Metadata, `Participant ID`, Group, Run, Gender, HIV_Status) %>%
             pivot_longer(Group:HIV_Status, names_to = "Variable", values_to = "Group"),
           by = c("Name" = "Participant ID"))

Temp %>%
  ggplot(aes(x = PC1, y = PC2, col = Group))+
    geom_point()+
    stat_ellipse(level = 0.95)+
    ggsci::scale_color_npg()+
    theme_bw()+
    facet_wrap(~Variable)+
    ggtitle("Principle Component Analysis Across Normalized Read Counts of Significant Gender Genes PC1: 58%, PC2: 6%")

summary(PCA)
```


# Run Variable PCA
```{r}
Group <- Out_New %>%
  filter(grepl("Group- Run", Contrast))

Filter <- Group$gene_id

Filter

#Create New DeSeq Obhect Using this for reference file:///Users/tporter/Downloads/RNASEQ20_Day3_HandsOn%20(1).pdf
PCA_Data <- DESeqDataSetFromMatrix(countData = TB_RNA,
                              colData = Metadata,
                              design = ~ Group + Run + Gender + HIV_Status)

PCA_Data <- estimateSizeFactors(PCA_Data)

sizeFactors(PCA_Data)

PCA_normalized <- vst(PCA_Data, blind = T)

# Convert tibble to data frame
PCA_normalized <- as.data.frame(head(assay(PCA_normalized), nrow(PCA_normalized)))

PCA_normalized$gene <- rownames(PCA_normalized)

PCA_normalized_Filtered <- PCA_normalized %>%
  filter(gene %in% Filter)

PCA_normalized_Filtered <- PCA_normalized_Filtered %>%
  pivot_longer(-gene, values_to = "Norm_Count", names_to = "Name") %>%
  pivot_wider(values_from = Norm_Count, names_from = gene)

PCA <- prcomp(PCA_normalized_Filtered[,-c(1)], center = T, scale = T)

PCA_Table <- data_frame("Name" = PCA_normalized_Filtered$Name, "PC1" = PCA$x[,1], "PC2" = PCA$x[,2])

PCA_Table

names(PCA_Table)

Temp <- right_join(PCA_Table,
           dplyr::select(Metadata, `Participant ID`, Group, Run, Gender, HIV_Status) %>%
             pivot_longer(Group:HIV_Status, names_to = "Variable", values_to = "Group"),
           by = c("Name" = "Participant ID"))

Temp %>%
  ggplot(aes(x = PC1, y = PC2, col = Group))+
    geom_point()+
    stat_ellipse(level = 0.95)+
    ggsci::scale_color_aaas()+
    theme_bw()+
    facet_wrap(~Variable)+
    ggtitle("Principle Component Analysis Across Normalized Read Counts of Significant Run Genes PC1: 39%, PC2: 12%")

summary(PCA)
```

# CASS Outcome PCA
```{r}

unique(Out_New$Contrast)

Group <- Out_New %>%
  filter(grepl("CASS Outcome- Negative vs. Positive", Contrast))

Filter <- Group$gene_id

Filter

#Create New DeSeq Obhect Using this for reference file:///Users/tporter/Downloads/RNASEQ20_Day3_HandsOn%20(1).pdf
PCA_Data <- DESeqDataSetFromMatrix(countData = TB_RNA,
                              colData = Metadata,
                              design = ~ CASS_outcome + Run + Gender + HIV_Status)

PCA_Data <- estimateSizeFactors(PCA_Data)

sizeFactors(PCA_Data)

PCA_normalized <- vst(PCA_Data, blind = T)

# Convert tibble to data frame
PCA_normalized <- as.data.frame(head(assay(PCA_normalized), nrow(PCA_normalized)))

PCA_normalized$gene <- rownames(PCA_normalized)

PCA_normalized_Filtered <- PCA_normalized %>%
  filter(gene %in% Filter)

PCA_normalized_Filtered <- PCA_normalized_Filtered %>%
  pivot_longer(-gene, values_to = "Norm_Count", names_to = "Name") %>%
  pivot_wider(values_from = Norm_Count, names_from = gene)

PCA <- prcomp(PCA_normalized_Filtered[,-c(1)], center = T, scale = T)

PCA_Table <- data_frame("Name" = PCA_normalized_Filtered$Name, "PC1" = PCA$x[,1], "PC2" = PCA$x[,2])

PCA_Table

names(PCA_Table)

Temp <- right_join(PCA_Table,
           dplyr::select(Metadata, `Participant ID`, CASS_outcome, Run, Gender, HIV_Status) %>%
             pivot_longer(CASS_outcome:HIV_Status, names_to = "Variable", values_to = "Group"),
           by = c("Name" = "Participant ID"))

Temp %>%
  ggplot(aes(x = PC1, y = PC2, col = Group))+
    geom_point()+
    stat_ellipse(level = 0.95)+
    ggsci::scale_color_jama()+
    theme_bw()+
    facet_wrap(~Variable)+
    ggtitle("Principle Component Analysis Across Normalized Read Counts of Significant Run Genes PC1: 30%, PC2: 11%")

summary(PCA)
```

# CASS Smear PCA
```{r}
#This creates a DESeq object
DeedaRNASeq_Culture <- DESeqDataSetFromMatrix(countData = TB_RNA_CASS_SMEAR,
                              colData = Metadata_CASS_SMEAR,
                              design = ~ CASS_Culture + Run + Gender + HIV_Status)


unique(Out_New$Contrast)

Group <- Out_New %>%
  filter(grepl("CASS Smear- Negative vs. Positive", Contrast))

Filter <- Group$gene_id

Filter

#Create New DeSeq Obhect Using this for reference file:///Users/tporter/Downloads/RNASEQ20_Day3_HandsOn%20(1).pdf
PCA_Data <- DESeqDataSetFromMatrix(countData = TB_RNA_CASS_SMEAR,
                              colData = Metadata_CASS_SMEAR,
                              design = ~ CASS_Smear + Run + Gender + HIV_Status)

PCA_Data <- estimateSizeFactors(PCA_Data)

sizeFactors(PCA_Data)

PCA_normalized <- vst(PCA_Data, blind = T)

# Convert tibble to data frame
PCA_normalized <- as.data.frame(head(assay(PCA_normalized), nrow(PCA_normalized)))

PCA_normalized$gene <- rownames(PCA_normalized)

PCA_normalized_Filtered <- PCA_normalized %>%
  filter(gene %in% Filter)

PCA_normalized_Filtered <- PCA_normalized_Filtered %>%
  pivot_longer(-gene, values_to = "Norm_Count", names_to = "Name") %>%
  pivot_wider(values_from = Norm_Count, names_from = gene)

PCA <- prcomp(PCA_normalized_Filtered[,-c(1)], center = T, scale = T)

PCA_Table <- data_frame("Name" = PCA_normalized_Filtered$Name, "PC1" = PCA$x[,1], "PC2" = PCA$x[,2])

PCA_Table

names(PCA_Table)

Temp <- right_join(PCA_Table,
           dplyr::select(Metadata, `Participant ID`, CASS_Smear, Run, Gender, HIV_Status) %>%
             pivot_longer(CASS_Smear:HIV_Status, names_to = "Variable", values_to = "Group"),
           by = c("Name" = "Participant ID"))

Temp %>%
  ggplot(aes(x = PC1, y = PC2, col = Group))+
    geom_point()+
    stat_ellipse(level = 0.95)+
    ggsci::scale_color_jama()+
    theme_bw()+
    facet_wrap(~Variable)+
    ggtitle("Principle Component Analysis Across Normalized Read Counts of Significant CASS Smear Genes PC1: 32%, PC2: 12%")

summary(PCA)
```

# CASS Culture PCA

```{r}
#This creates a DESeq object
DeedaRNASeq_Culture <- DESeqDataSetFromMatrix(countData = TB_RNA_CASS_SMEAR,
                              colData = Metadata_CASS_SMEAR,
                              design = ~ CASS_Culture + Run + Gender + HIV_Status)


unique(Out_New$Contrast)

Group <- Out_New %>%
  filter(grepl("CASS Smear- Negative vs. Positive", Contrast))

Filter <- Group$gene_id

Filter

#Create New DeSeq Obhect Using this for reference file:///Users/tporter/Downloads/RNASEQ20_Day3_HandsOn%20(1).pdf
PCA_Data <- DESeqDataSetFromMatrix(countData = TB_RNA_CASS_SMEAR,
                              colData = Metadata_CASS_SMEAR,
                              design = ~ CASS_Culture + Run + Gender + HIV_Status)

PCA_Data <- estimateSizeFactors(PCA_Data)

sizeFactors(PCA_Data)

PCA_normalized <- vst(PCA_Data, blind = T)

# Convert tibble to data frame
PCA_normalized <- as.data.frame(head(assay(PCA_normalized), nrow(PCA_normalized)))

PCA_normalized$gene <- rownames(PCA_normalized)

PCA_normalized_Filtered <- PCA_normalized %>%
  filter(gene %in% Filter)

PCA_normalized_Filtered <- PCA_normalized_Filtered %>%
  pivot_longer(-gene, values_to = "Norm_Count", names_to = "Name") %>%
  pivot_wider(values_from = Norm_Count, names_from = gene)

PCA <- prcomp(PCA_normalized_Filtered[,-c(1)], center = T, scale = T)

PCA_Table <- data_frame("Name" = PCA_normalized_Filtered$Name, "PC1" = PCA$x[,1], "PC2" = PCA$x[,2])

PCA_Table

names(PCA_Table)

Temp <- right_join(PCA_Table,
           dplyr::select(Metadata, `Participant ID`, CASS_Culture, Run, Gender, HIV_Status) %>%
             pivot_longer(CASS_Culture:HIV_Status, names_to = "Variable", values_to = "Group"),
           by = c("Name" = "Participant ID"))

Temp %>%
  ggplot(aes(x = PC1, y = PC2, col = Group))+
    geom_point()+
    stat_ellipse(level = 0.95)+
    ggsci::scale_color_jama()+
    theme_bw()+
    facet_wrap(~Variable)+
    ggtitle("Principle Component Analysis Across Normalized Read Counts of Significant CASS Culture Genes PC1: 32%, PC2: 12%")

summary(PCA)
```


# Functional Analysis Using GAGE
```{r}
BiocManager::install("gage")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("AnnotationDbi")

library("AnnotationDbi")
library("org.Hs.eg.db")
library("gage")

sessionInfo()


res <- results(DeedaRNASeq_Analysis_Culture, contrast=c("HIV_Status","Positive", "Negative"), alpha=0.05)

#res$symbol = 
AnnotationDbi::mapIds(org.Hs.eg.db,
                     keys=row.names(res), 
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")

```































# Functional Analysis
```{r}
mart <- useMart('ENSEMBL_MART_ENSEMBL')
mart <- useDataset('hsapiens_gene_ensembl', mart)

mart <- useEnsembl(biomart = "ensembl", 
                     dataset = "hsapiens_gene_ensembl")

res <- results(DeedaRNASeq_Analysis_CASS_outcome, contrast=c("HIV_Status","Negative", "Positive"), alpha=0.05)

all_genes <- as.character(Out_New$ensembl_gene_id)

genes <- getBM(filters="ensembl_gene_id",
 attributes=c("ensembl_gene_id", "entrezgene_id"),
 values= all_genes,
 mart=mart)

indNA = which(is.na(genes$entrezgene_id))
genes_noNA <- genes[-indNA,]
indnodup = which(duplicated(genes_noNA$ entrezgene_id) == F)
genes_noNA_nodup <- genes_noNA[indnodup,]
lFC <- res$log2FoldChange[-indNA]
lFC <- lFC[indnodup]
names(lFC) <- genes_noNA_nodup$entrezgene_id
# Sort fold changes in decreasing order
lFC <- sort(lFC, decreasing = TRUE)

save(lFC, file='/scratch/tporter/TB_20231117_Megans_DESeq/Deeda_RNA_Seq_02.24_lFC.RData')
```


```{r}
load('/Volumes/scratch/tporter/TB_20231117_Megans_DESeq/Deeda_RNA_Seq_02.24_lFC.RData')

BiocManager::install("clusterProfiler", force = T)

ngseaKEGG <- clusterProfiler::gseKEGG(geneList = lFC,
 organism = "hsa",
 nPerm = 1000, # default number permutations
 minGSSize = 5, # minimum gene set size
 pvalueCutoff = 0.1, # padj cutoff value
 verbose = FALSE)
# Extract the GSEA results
gseaKEGG_results <- gseaKEGG@result


??clusterProfiler::gseKEGG
```

Group <- data_frame(TB_RNA_DF) %>%
  filter(Gene %in% Filter)

Group_Wide <- Group %>%
  dplyr::select(Gene, Name, Gender, Reads) %>%
  pivot_wider(values_from = Reads, names_from = Gene)

Group_Wide[,-c(1:2)]

PCA <- prcomp(Group_Wide[,-c(1:2)], center = T, scale = T)

summary(PCA)

Temp <- cbind(Group_Wide[,c(1:2)], PCA$x) 

Plotly <- Temp %>%
  ggplot(aes(x = PC1, y = PC2, col = Gender, group = Name))+
    geom_point()+
    ggsci::scale_color_nejm()+
    theme_bw()+
    ggtitle("PCA of significant genes within the Gender contrast")

plotly::ggplotly(Plotly)

htmlwidgets::saveWidget(plotly::as_widget(plotly::ggplotly(Plotly, width = 1200, height = 800)),
                        file="/scratch/tporter/TB_20231117_Megans_DESeq/PCA_Gender.html", selfcontained = TRUE)


library(ggbiplot)
ggbiplot(PCA,
              obs.scale = 1,
              var.scale = 1,
              groups = Temp$Gender,
              ellipse = TRUE,
              circle = TRUE,
              ellipse.prob = 0.68,
              var.axes = F)+
  scale_color_discrete(name = '')+
  theme(legend.direction = 'horizontal',
               legend.position = 'top')+
  theme_bw()+
  ggsci::scale_color_nejm()+
  ggtitle("PCA of significant genes within the HIV contrast")
